---
title: "test"
author: "AlexBender"
date: "November 28, 2018"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```


We will load that data into a sparse matrix with 70 columns instead of a data frame using the arules package. 

```{r}
#install.packages("arules")
library(arules)
```
The way the data is formatted, it wouldn't be correct to just import as a dataframe. As the data description states, it is in transactional form, so we are going to use the read.transactions function from the arules package, which will read in the data correctly. We should see a sparse matrix with 70 columns and ~34000 rows. The read transactions function creates an object of class transactions in order to load information from the data. 
```{r}
#load the data as a spare matrix
plants <- read.transactions("plants.data.txt", format = "basket", sep = ",", cols = 1)
#explore data
summary(plants)
```
This summary shows us good information about the plant "transactions". We can see that California is the most common location for the plants to be located. Additionally, there is an average of ~9 distinct states/provinces that a plant is located in. 

Let's explore a little more. 

```{r}
#explore the first 5 transaction sets
inspect(plants[1:5])
```
This is broken up just like the original CSV file, but the format is a bit weird. Let's explore a more typical format. 

```{r}
#explore frequency of a few states
itemFrequency(plants[ , 1:3])
```
Now let's explore these transactions visually. Here are the states/provinces with at least 15% support. 

```{r}
#plot states with more than 15% support
itemFrequencyPlot(plants, support = 0.15)
```

Now let's plot the top 15 states/provinces. 

```{r}
#plot top 15 states
itemFrequencyPlot(plants, topN = 15)
```
Let's visualize the spare matrix using the image() function. 

```{r}
image(plants[1:10])
```

The number of small gray squares per row indicates the number of states per plant. 

It's hard to visualize a lot of rows in this, but let's try to visualize random samples. 

```{r}
image(sample(plants, 75))
```

Now that we have an idea of the way this data is loaded, let's convert this to a useable form for k-means clustering. Since kmeans uses distance measues and coordinate centroids to determine clusters, all features must be numerical. Currently we have only character data. 

To do this, we are going to convert the transactional sparse matrix into a logical matrix. The columns will be the species name and then the 69 different states and provinces located in the data -- one for each state/province. The rows will be each species followed by a FALSE in the columns in which that species is not located and a TRUE in the columns in which the species is located. 

```{r}
#the arules function as() with the "matrix" parameter converts the transactional sparse matrix into a logical matrix
plant_matrix <- as(plants, "matrix")
#let's make sure it worked
head(plant_matrix[,1:5])
```

As you can see, this matrix contains the 70 columns in the data with a TRUE in states that the species is located and a FALSE in the states that the species is not located. 

We still can't do clustering with this since it is not numerical. Our end goal will be this same matrix but with a 1 for TRUE (presence) and a 0 for FALSE (absence). In order to convert this to numerical we are going to do a very simple solution: multiple the whole matrix by 1. If a cell = TRUE, 1*1=1. If a cell = FALSE, 1*0=0.
```{r}
#multiple matrix by 1
plant_matrix <- 1*plant_matrix
#see if it worked
head(plant_matrix[,1:5])
```
Perfect! To do a quick spot check, based on the data we know that the 5th species (abies) is located in Alabama(ab), Alaska(ak), and Arizona(az), but not Alberta(al) or Arkansas(ar).  This checks out because the only columns with 1s in them are ab, ak, az for abies. 

Now that we have our binary/numerical matrix of species locations, let's do some clustering! 

We know that a Hamming distance is great for binary distances. Since all values are either 0 or 1, we could use a hamming distance for the k-means clustering. The default for the kmeans package is Euclidean, however. 

First, let's explore if there are any NA values. 

```{r}
#explore if NAs
summary(plant_matrix)
```
We are assuming that all 1 values were inputted correctly and any missing values turned into 0s, meaning that if it is missing data this would be marked as absence of this species in this area. There are no additional NA values, woo!

However, if there are species that are not present in any of the states/provinces, then we don't need them in our clustering since they don't add any value. We are going to check for this by taking a sum of the rows. If the sum is 0, meaning all values are 0, we will remove that row. 

```{r}
#display which rows have a 0 sum
plant_matrix[which(rowSums(plant_matrix)==0)]
```

There are no species that have no locations; this is good!


Now let's convert the matrix to a data frame in order to properly cluster the data. 

```{r}
plant_df <- as.data.frame(plant_matrix)
```

Choose k by geographical/climate regions as well as by plant hieracrhy

Choosing a proper value for k is one of the most difficult things in k-means clustering. Since there is no hard and fast rule, it is a kind of arbitrary process. The best way to pick a k value is through subject matter expertise and context knowledge. For this example, three possible cluster amounts come to mind. First, we can cluster based on geographical regions. Second, we can cluster based on climactic regions/habitats. Third, we can cluster based on classification families of the plants located within the data.

First, we are going to use a k of 8, which represents the 8 major physical/geographic regions in North America: 

1) Coastal Plains
2) Appalachian Mountains
3) Canadian Shield
4) Interior Lowlands
5) Great Plains
6) Rocky Mountains
7) Basin and Range
8) Coastal Range

```{r}
#set random seed to follow
set.seed(1234)
plant_clusters <- kmeans(plant_df, 8)
```


Now we have our clusters! Let's investigate further and evaluate performance. 

Let's check out the relative size of the clusters.

```{r}
plant_clusters$size
```

There is a pretty large disparity is cluster size. Let's check out the cluster centroids now. 

```{r}
plant_clusters$centers
```


If a number here is above 0.5 it indicates that the cluster is above average in interest in that category.The closer to 1 the stronger the relationship between the state and the cluster. The closer to 0, the weaker the relationship. For example, clusters 4, 5, 6 are above average in Wyoming (and other states as well). 

Now let's try try to improve model performance. 

Let's add the clusters as a column to the original data frame. 

```{r}
plant_df$cluster <- plant_clusters$cluster
```

Check it out with some states in the SOuthern USA.

```{r}
plant_df[1:5, c("cluster", "al", "ar", "la", "ms", "fl")]
```
As we can see, some of the species that are prevalent in similar geographic zones are indeed in the same cluster.

Let's see the states/provinces represented by each cluster. We are going to do this by adding up all of the 1s located in each state's column.

```{r}
library(tidyverse)
#filter the data by cluster number then add columns and transpose
clust1_top <- plant_df %>% filter(cluster == 1) %>% summarise_all(funs(sum)) %>% t()
#remove cluster row and make data frame
clust1_top <- as.data.frame(clust1_top[-71,])
#rename sum column as number of plants
colnames(clust1_top) <- "numberOfPlants"
#make a 1st column with the state names
clust1_top <- rownames_to_column(clust1_top, "State/Province")
#put columns in descending order by number of plants
clust1_top <- clust1_top[order(-clust1_top$numberOfPlants),]
#display top 10 by cluster
clust1_top[1:10,]
```
The purpose of printing out the top 10 states is to illustrate the geographic regions in which the relevant cluster of plants is primarily located. In this case, it seems that it may have been clustered into the Basin and Range region (CA, TX, AZ, NM, OR, NV, UT, WA, CO)


```{r}
#filter the data by cluster number then add columns and transpose
clust2_top <- plant_df %>% filter(cluster == 2) %>% summarise_all(funs(sum)) %>% t()
#remove cluster row and make data frame
clust2_top <- as.data.frame(clust2_top[-71,])
#rename sum column as number of plants
colnames(clust2_top) <- "numberOfPlants"
#make a 1st column with the state names
clust2_top <- rownames_to_column(clust2_top, "State/Province")
#put columns in descending order by number of plants
clust2_top <- clust2_top[order(-clust2_top$numberOfPlants),]
#display top 10 by cluster
clust2_top[1:10,]
```
This cluster seems to represent the Appalachian Mts. region (NY, MA, PA, CT, ON, ME, NJ, NH, VT)

```{r}
#filter the data by cluster number then add columns and transpose
clust3_top <- plant_df %>% filter(cluster == 3) %>% summarise_all(funs(sum)) %>% t()
#remove cluster row and make data frame
clust3_top <- as.data.frame(clust3_top[-71,])
#rename sum column as number of plants
colnames(clust3_top) <- "numberOfPlants"
#make a 1st column with the state names
clust3_top <- rownames_to_column(clust3_top, "State/Province")
#put columns in descending order by number of plants
clust3_top <- clust3_top[order(-clust3_top$numberOfPlants),]
#display top 10 by cluster
clust3_top[1:10,]
```
This cluster seems to represent the Coastal Plain region (PR, VI, FL, TX, LA, GA, AL)

```{r}
#filter the data by cluster number then add columns and transpose
clust4_top <- plant_df %>% filter(cluster == 4) %>% summarise_all(funs(sum)) %>% t()
#remove cluster row and make data frame
clust4_top <- as.data.frame(clust4_top[-71,])
#rename sum column as number of plants
colnames(clust4_top) <- "numberOfPlants"
#make a 1st column with the state names
clust4_top <- rownames_to_column(clust4_top, "State/Province")
#put columns in descending order by number of plants
clust4_top <- clust4_top[order(-clust4_top$numberOfPlants),]
#display top 10 by cluster
clust4_top[1:10,]
```
This cluster seems to represent the Interior Lowlands region (IL, NY, PA, OH, VA, MI, WI, ON, IN, MO)

```{r}
#filter the data by cluster number then add columns and transpose
clust5_top <- plant_df %>% filter(cluster == 5) %>% summarise_all(funs(sum)) %>% t()
#remove cluster row and make data frame
clust5_top <- as.data.frame(clust5_top[-71,])
#rename sum column as number of plants
colnames(clust5_top) <- "numberOfPlants"
#make a 1st column with the state names
clust5_top <- rownames_to_column(clust5_top, "State/Province")
#put columns in descending order by number of plants
clust5_top <- clust5_top[order(-clust5_top$numberOfPlants),]
#display top 10 by cluster
clust5_top[1:10,]
```
This cluster seems to represent the Canadian Shield region (ON, BC, QC, AB, MB, SK, MN, MI, WI)

```{r}
#filter the data by cluster number then add columns and transpose
clust6_top <- plant_df %>% filter(cluster == 6) %>% summarise_all(funs(sum)) %>% t()
#remove cluster row and make data frame
clust6_top <- as.data.frame(clust6_top[-71,])
#rename sum column as number of plants
colnames(clust6_top) <- "numberOfPlants"
#make a 1st column with the state names
clust6_top <- rownames_to_column(clust6_top, "State/Province")
#put columns in descending order by number of plants
clust6_top <- clust6_top[order(-clust6_top$numberOfPlants),]
#display top 10 by cluster
clust6_top[1:10,]
```
This cluster seems to represent the Rocky Mts. region (ID, MT, WY, UT, CO, BC)

```{r}
#filter the data by cluster number then add columns and transpose
clust7_top <- plant_df %>% filter(cluster == 7) %>% summarise_all(funs(sum)) %>% t()
#remove cluster row and make data frame
clust7_top <- as.data.frame(clust7_top[-71,])
#rename sum column as number of plants
colnames(clust7_top) <- "numberOfPlants"
#make a 1st column with the state names
clust7_top <- rownames_to_column(clust7_top, "State/Province")
#put columns in descending order by number of plants
clust7_top <- clust7_top[order(-clust7_top$numberOfPlants),]
#display top 10 by cluster
clust7_top[1:10,]
```
This cluster seems to represent either the Appalachian Mts. or Coastal Plain regions.

```{r}
#filter the data by cluster number then add columns and transpose
clust8_top <- plant_df %>% filter(cluster == 8) %>% summarise_all(funs(sum)) %>% t()
#remove cluster row and make data frame
clust8_top <- as.data.frame(clust8_top[-71,])
#rename sum column as number of plants
colnames(clust8_top) <- "numberOfPlants"
#make a 1st column with the state names
clust8_top <- rownames_to_column(clust8_top, "State/Province")
#put columns in descending order by number of plants
clust8_top <- clust8_top[order(-clust8_top$numberOfPlants),]
#display top 10 by cluster
clust8_top[1:10,]
```
This cluster seems to represent either the Interior Lowlands or Coastal Plain regions.

This k-means clustering shows significant actionable variation between the 8 clusters. 

Next, we are going to use a k of 10, which represents the 10 climactic regions/habitat types of North America: 

1) Tropical Moist Broadleaf Forests
2) Tropical Dry Broadleaf Forests
3) Temperate Broadleaf and Mixed Forests
4) Temperate Coniferous Forests
5) Temperate Grasslands, Savannas, and Shrublands
6) Flooded Grasslands
7) Mediterranean Forests, Woodlands, Savanna, and Scrub
8) Deserts and Xeric Shrublands
9) Taiga and Boreal Forests
10) Tundra 

```{r}
#set random seed to follow
set.seed(1234)
plant_clusters10 <- kmeans(plant_df, 10)
```

Now we have our clusters! Let's investigate further and evaluate performance. 

Let's check out the relative size of the clusters.

```{r}
plant_clusters10$size
```
As expected, some of the clusters have changed in size from the previous iteration, indicating that there should be some differences in the clustering. Let's take a look.

```{r}
#add the clusters10 as a column to the original data frame
plant_df$cluster10 <- plant_clusters10$cluster
```

Check it out with some states in the Southeastern USA.

```{r}
plant_df[1:5, c("cluster10", "fl", "ga", "nc", "sc")]
```

Once again, let's see the states/provinces represented by each cluster.
```{r}
#filter the data by cluster number then add columns and transpose
clust1_top1 <- plant_df %>% filter(cluster10 == 1) %>% summarise_all(funs(sum)) %>% t()
#remove cluster row and make data frame
clust1_top1 <- as.data.frame(clust1_top1[-c(71,72),])
#rename sum column as number of plants
colnames(clust1_top1) <- "numberOfPlants"
#make a 1st column with the state names
clust1_top1 <- rownames_to_column(clust1_top1, "State/Province")
#put columns in descending order by number of plants
clust1_top1 <- clust1_top1[order(-clust1_top1$numberOfPlants),]
#display top 10 by cluster
clust1_top1[1:10,]
```
This cluster seems to represent the Deserts and Xeric Shrublands region (CA, OR, TX, WA, NV, ID)

```{r}
#filter the data by cluster number then add columns and transpose
clust2_top1 <- plant_df %>% filter(cluster10 == 2) %>% summarise_all(funs(sum)) %>% t()
#remove cluster row and make data frame
clust2_top1 <- as.data.frame(clust2_top1[-c(71,72),])
#rename sum column as number of plants
colnames(clust2_top1) <- "numberOfPlants"
#make a 1st column with the state names
clust2_top1 <- rownames_to_column(clust2_top1, "State/Province")
#put columns in descending order by number of plants
clust2_top1 <- clust2_top1[order(-clust2_top1$numberOfPlants),]
#display top 10 by cluster
clust2_top1[1:10,]
```
This cluster seems to represent the Temperate Broadleaf and Mixed Forests region (NY, MA, PA, CT, ON, ME, MI, NJ, NH, VT)

```{r}
#filter the data by cluster number then add columns and transpose
clust3_top1 <- plant_df %>% filter(cluster10 == 3) %>% summarise_all(funs(sum)) %>% t()
#remove cluster row and make data frame
clust3_top1 <- as.data.frame(clust3_top1[-c(71,72),])
#rename sum column as number of plants
colnames(clust3_top1) <- "numberOfPlants"
#make a 1st column with the state names
clust3_top1 <- rownames_to_column(clust3_top1, "State/Province")
#put columns in descending order by number of plants
clust3_top1 <- clust3_top1[order(-clust3_top1$numberOfPlants),]
#display top 10 by cluster
clust3_top1[1:10,]
```
This cluster seems to represent a mix of various climates -- including Tropical Moist Broadleaf Forests, Flooded Grasslands, and Temperate Grasslands

```{r}
#filter the data by cluster number then add columns and transpose
clust4_top1 <- plant_df %>% filter(cluster10 == 4) %>% summarise_all(funs(sum)) %>% t()
#remove cluster row and make data frame
clust4_top1 <- as.data.frame(clust4_top1[-c(71,72),])
#rename sum column as number of plants
colnames(clust4_top1) <- "numberOfPlants"
#make a 1st column with the state names
clust4_top1 <- rownames_to_column(clust4_top1, "State/Province")
#put columns in descending order by number of plants
clust4_top1 <- clust4_top1[order(-clust4_top1$numberOfPlants),]
#display top 10 by cluster
clust4_top1[1:10,]
```
This cluster seems to represent the Tempereate Broadleaf and Mixed Forests region (PA, NY, VA, MD, OH, TN, KY, IL, CT, NJ)

```{r}
#filter the data by cluster number then add columns and transpose
clust5_top1 <- plant_df %>% filter(cluster10 == 5) %>% summarise_all(funs(sum)) %>% t()
#remove cluster row and make data frame
clust5_top1 <- as.data.frame(clust5_top1[-c(71,72),])
#rename sum column as number of plants
colnames(clust5_top1) <- "numberOfPlants"
#make a 1st column with the state names
clust5_top1 <- rownames_to_column(clust5_top1, "State/Province")
#put columns in descending order by number of plants
clust5_top1 <- clust5_top1[order(-clust5_top1$numberOfPlants),]
#display top 10 by cluster
clust5_top1[1:10,]
```
This cluster seems to represent the Taija and Boreal Forests region (ON, BC, QC, AB, MB, SK, MN)

```{r}
#filter the data by cluster number then add columns and transpose
clust6_top1 <- plant_df %>% filter(cluster10 == 6) %>% summarise_all(funs(sum)) %>% t()
#remove cluster row and make data frame
clust6_top1 <- as.data.frame(clust6_top1[-c(71,72),])
#rename sum column as number of plants
colnames(clust6_top1) <- "numberOfPlants"
#make a 1st column with the state names
clust6_top1 <- rownames_to_column(clust6_top1, "State/Province")
#put columns in descending order by number of plants
clust6_top1 <- clust6_top1[order(-clust6_top1$numberOfPlants),]
#display top 10 by cluster
clust6_top1[1:10,]
```
This cluster seems to represent either the Deserts and Xeric Shrublands region or Temperate Grasslands, Savannas, and Shrublands region

```{r}
#filter the data by cluster number then add columns and transpose
clust7_top1 <- plant_df %>% filter(cluster10 == 7) %>% summarise_all(funs(sum)) %>% t()
#remove cluster row and make data frame
clust7_top1 <- as.data.frame(clust7_top1[-c(71,72),])
#rename sum column as number of plants
colnames(clust7_top1) <- "numberOfPlants"
#make a 1st column with the state names
clust7_top1 <- rownames_to_column(clust7_top1, "State/Province")
#put columns in descending order by number of plants
clust7_top1 <- clust7_top1[order(-clust7_top1$numberOfPlants),]
#display top 10 by cluster
clust7_top1[1:10,]
```
This cluster seems to represent either the Temperate Broadleaf and Mixed Forests region or Temperate Coniferous Forests region

```{r}
#filter the data by cluster number then add columns and transpose
clust8_top1 <- plant_df %>% filter(cluster10 == 8) %>% summarise_all(funs(sum)) %>% t()
#remove cluster row and make data frame
clust8_top1 <- as.data.frame(clust8_top1[-c(71,72),])
#rename sum column as number of plants
colnames(clust8_top1) <- "numberOfPlants"
#make a 1st column with the state names
clust8_top1 <- rownames_to_column(clust8_top1, "State/Province")
#put columns in descending order by number of plants
clust8_top1 <- clust8_top1[order(-clust8_top1$numberOfPlants),]
#display top 10 by cluster
clust8_top1[1:10,]
```
This cluster seems to represent either the Temperate Broadleaf and Mixed Forests region or Temperate Coniferous Forests region

```{r}
#filter the data by cluster number then add columns and transpose
clust9_top <- plant_df %>% filter(cluster10 == 9) %>% summarise_all(funs(sum)) %>% t()
#remove cluster row and make data frame
clust9_top <- as.data.frame(clust9_top[-c(71,72),])
#rename sum column as number of plants
colnames(clust9_top) <- "numberOfPlants"
#make a 1st column with the state names
clust9_top <- rownames_to_column(clust9_top, "State/Province")
#put columns in descending order by number of plants
clust9_top <- clust9_top[order(-clust9_top$numberOfPlants),]
#display top 10 by cluster
clust9_top[1:10,]
```
This cluster seems to represent the Temperate Coniferous Forests (ID, MT, WY, UT, OR, CO, WA, CA, BC)

```{r}
#filter the data by cluster number then add columns and transpose
clust10_top <- plant_df %>% filter(cluster10 == 10) %>% summarise_all(funs(sum)) %>% t()
#remove cluster row and make data frame
clust10_top <- as.data.frame(clust10_top[-c(71,72),])
#rename sum column as number of plants
colnames(clust10_top) <- "numberOfPlants"
#make a 1st column with the state names
clust10_top <- rownames_to_column(clust10_top, "State/Province")
#put columns in descending order by number of plants
clust10_top <- clust10_top[order(-clust10_top$numberOfPlants),]
#display top 10 by cluster
clust10_top[1:10,]
```
This cluster seems to represent the Temperate Broadleaf and Mixed Forests region (IL, NY, PA, OH, VA, MI, WI, ON, IN, MO)

Finally,  we would consider trying to cluster by plant classification hierarchy. Because it is specified that the names provided in the dataset is either the plant genus or species, we will go one level above in the hierarchy (Family) in order to determine an appropriate k value. After surveying several plant families, it seemed that most families contained about 10-20 genera within each (i.e. Sapindales (10), Cupressaceae (15), Polypodiales (20)). As a result, we will set our k to 15 in order to attempt to classify based on plant classification. Disclaimer: we would need t consult a plant expert in order to properly choose the value of k. 15 is our assumption.

```{r}
#set random seed to follow
set.seed(1234)
plant_clusters15 <- kmeans(plant_df, 15)
```

Now we have our clusters! Let's investigate further and evaluate performance. 

Let's check out the relative size of the clusters.

```{r}
plant_clusters15$size
```

```{r}
#add the clusters15 as a column to the original data frame
plant_df$cluster15 <- plant_clusters15$cluster
```

Check it out with some states in the Southeastern USA.

```{r}
#not sure which columns to include here... i guess maybe just cluster20? but seems weird to not compare to any base metrics
plant_df[1:10, c("cluster15", "cluster10","fl", "ga", "nc", "sc")]
```
By clustering by family, we have already made the underlying assumptio that Kingdom = Plantae and Subkindom = Vascular plants. However, since we are not plant experts, we cannot necessarily draw conclusions from this iteration of clustering and therefore cannot match up each cluster to a specific family, but would now hand off this part of the analysis to a plant expert.



Theoretically, we would try to run our clustering approaches again, but using Hamming (binary) distance instead of Euclidean. However, we weren't able to find a package to implement this seamlessly.


Now let's run with our analysis using 8 and 10 clusters based on geographic and climactic regions. We are going to attempt to visualize these clusters. To do this we will be using the packages factoextra and cluster. 
```{r}
#if(!require(devtools)) install.packages("devtools")
#devtools::install_github("kassambara/factoextra")
#install.packages("cluster")
library(factoextra)
library(cluster)
```

Let's create a visualization of the 8 clusters. 
```{r}
#visualize clusters minus added cluster columns
fviz_cluster(plant_clusters, data = plant_df[-c(71:73)], geom = "point", stand = FALSE, ellipse.type = "norm")
```
This plot shows clustering based on Dim1 and Dim2, which are the principal components. 

With 8 clusters it seems as though there is a decent mount of crossover between clusters. This makes sense because the grouping based on geographic region is quite arbitrary and a state/province border is an arbitrary line that wouldn't stop plants from crossing over. Clustering based on climactic/habitat regions may be more logical.

Let's create a visualization of the 10 clusters of habitats/climates. 
```{r}
#visualize clusters minus added cluster columns
fviz_cluster(plant_clusters10, data = plant_df[-c(71:73)], geom = "point", stand = FALSE, ellipse.type = "norm")
```

Here there is also a good amount of overlap, but this makes sense. Every state is located within every cluster because there may be a small amount of plant species from a cluster located in states in which they are not very prevalent. Because of this representation of every state/province in every cluster, you would expect them to overlap. However, there are also significant portions of each cluster that do not overlap (such as the large greenish section near the bottom). These portions indicate the parts of the cluster that are unique- ideally the different climt/habitat regions. This visualization is here to help us understand how the cluster vary across the principal components.


Now let's do an elbow plot to determine a possible optimal value of k mathematically based on homogeneity within the clusters at different k values. We will see how our context/subject matter guesses pan out. 
```{r}
set.seed(1234)
# Compute and plot wss for k = 2 to k = 15
# Maximal number of clusters based on when the graph inflects
k.max <- 20 
#calculate within cluster homogeneity at each value of k
wss <- sapply(1:k.max, function(k){kmeans(plant_df[-c(71:73)], k)$tot.withinss})
#plot the homogenity "score" at each value of k 1:20. 
plot(1:k.max, wss,
       type="b", pch = 19, frame = FALSE, 
       xlab="Number of clusters K",
       ylab="Total within-clusters sum of squares")
#plot a line at the supposed inflection point
abline(v = 7, lty =2)
```
It looks like anywhere in the range of 5-10 seems like a good number of clusters; this is pretty spot on with our initial educated guesses.

This amount of clusters (5-10) balances not overfitting while still having high homogeneity within clusters. 

**Are there clusters in the data?** Yes, as analyzed above!


For fun, let's run the apriori algorithm to see if we can find any non-trival association rules!

```{r}
plantrules <- apriori(plants)
```

We have a set of 506 rules. This is because by default we are only accepting the states with at least 3478 plants (10%). A lot of *hopefully* generalizeable rules were made. Let's see how useful these are by evaluating our model. 

```{r}
summary(plantrules)
```
This summary tells us a lot about our plant/state association rules. It seems that we chose good support and confidence levels since the lift is definitely high (>1)

Let's inspect a few rules.


```{r}
inspect(plantrules[2:4])
```

This shows us that 81% of plants that appear in Wyoming also appear in Montana. With a support of 0.11 and a lift of 5.8. This means that the data shows that there is a relationship between these two states having similar plants that is more likely than by chance alone. 

Now let's try to improve the model performance. Let's sort the rules by lift to see the most relevant ones. 

```{r}
inspect(sort(plantrules, by = "lift")[1:5])
```
These rules show that for example, plants in Connecticut are about 7 times more likely to also occur in Massachussets, New Jersey, and New York than the average state. 

Let's explore just the state: Oregon. 

```{r}
oregonrules <- subset(plantrules, items %in% "or")
inspect(oregonrules)
```

This is cool! We can see all states' flora associated with Oregon such as Washington, Idaho, and British Colombia. This is exactly what we'd expect based on Geography/climate. 

Now let's write these rules to a file. 

```{r}
write(plantrules, file = "plantrules.csv", sep = ",", quote = TRUE, row.names = FALSE)
```

Let's also try converting the rules to a dataframe. 

```{r}
plantrules_df <- as(plantrules, "data.frame")
#explore df 
str(plantrules_df)
summary(plantrules_df)
```

**References**

[1] 	STHDA, "Determining the optimal number of clusters: 3 must known methods - Unsupervised Machine Learning," Statistical Tools for High-Throughput Data Analysis, [Online]. Available: http://www.sthda.com/english/wiki/print.php?id=239. [Accessed 26 November 2018].

[2] 	Cephas, "Major habitat type CAN USA," Wikimedia Commons, 5 September 2010. [Online]. Available: https://commons.wikimedia.org/wiki/File:Major_habitat_type_CAN_USA.svg. [Accessed 26 November 2018].

[3] 	"Data Wrangling with dplyr and tidyr," RStudio, January 2015. [Online]. Available: https://www.rstudio.com/wp-content/uploads/2015/02/data-wrangling-cheatsheet.pdf. [Accessed 25 November 2018].

[4] 	M.-L. Liberatore, "8 Geographic Regions of North America," thinglink, November 2016. [Online]. Available: https://www.thinglink.com/scene/839995544494407684. [Accessed 27 November 2018].

[5] 	United States Department of Agriculture, "PLANTS Database," Natural Resources Conservation Service, 26 November 2018. [Online]. Available: https://plants.sc.egov.usda.gov/java/. [Accessed 26 November 2018].



